A Spectral Analysis of the Domain Decomposed Adjoint Neumann-Ulam Method

### 2012-11-30 doc/mandc2013/notes/spectral_theory.tex
paper updates for mandc theory

\subsection{Domain Leakage Approximations}
In a domain decomposed situation, not all histories will remain
within the domain they started in and must instead be
communicated. This communication, expected to be expensive, was
analyzed by Siegel and colleagues for idealized, load balanced
situations for full-core Monte Carlo\cite{siegel_analysis_2012}.  To
quantify the number of particles that leak out of the local domain
they define a leakage fraction, $\Lambda$, as:
\begin{equation}
  \Lambda = \frac{average\ \#\ of\ particles\ leaving\ local\ domain}
          {total\ of\ \#\ of\ particles\ starting\ in\ local\ domain}\:.
  \label{eq:leakage_fraction}
\end{equation}
For their studies, they assumed that the value of $\Lambda$ was
dependent on the total cross section of the system via the Wigner
rational approximation. This approximation was developed by Wigner to
compute the neutron leakage probability from the fuel in a reactor as
a simple interpolation between the small and large fuel lump cases,
approximately characterizing the leakage behavior of many common
geometries \cite{duderstadt_nuclear_1976}.

@@

In the case of domain decomposed linear operator equations, we can use
diffusion theory to estimate the optical thickness of a region and the
corresponding leakage fraction in terms of properties of the linear
operator and the discretization. To begin we must first calculate the
mean distance a Monte Carlo history will move in the grid by computing
the mean squared distance of its movement along the chord defined
across the domain of length $l$. After a single transition a history
will have moved a mean squared distance of:
\begin{equation}
  \langle \bar{r_1^2} \rangle = (n_s h)^2\:,
  \label{eq:step_1_length}
\end{equation}
where $h$ is the size of the grid along the chord and $n_s$ is the
number of grid elements a history will move on average every
transition. For our diffusion model problem, $n_s$ would equate to the
number of states in the $i$ (or $j$ as the problem is symmetric)
direction that a history could potentially move in a single transition
and is dependent on the stencil used for the discretization. After $k$
transitions in the random walk, the history will have moved a mean
squared distance of:
\begin{equation}
  \langle \bar{r_k^2} \rangle = k (n_s h)^2\:.
  \label{eq:step_k_length}
\end{equation}
If our our chord is of length $l$ and there are $n_i$ grid points (or
states to which a history may transition) along that chord, then $h =
l / n_i$ giving:
\begin{equation}
  \langle \bar{r_k^2} \rangle = k \Bigg(\frac{n_s l}{n_i}\Bigg)^2\:.
  \label{eq:step_k_length_sub}
\end{equation}
From diffusion theory, we expect the average number of interactions
along the chord to be:
\begin{equation}
  \tau = \frac{l}{2 d \sqrt{\langle \bar{r_k^2} \rangle}}\:,
  \label{eq:optical_thickness_1}
\end{equation}
where $d$ is the dimension of the problem and $\sqrt{\langle
  \bar{r_k^2} \rangle}$ is effectively the mean free path of the Monte
Carlo history in the domain. We can readily interperet $\tau$ to be
the \textit{effective optical thickness} of the domain of length
$l$. Inserting Eq~(\ref{eq:step_k_length_sub}) we arrive at:
\begin{equation}
  \tau = \frac{n_i}{2 d n_s \sqrt{k}}\:,
  \label{eq:optical_thickness_2}
\end{equation}
which if expanded with Eq~(\ref{eq:analytic_k}) gives us the final
relation for the effective optical thickness:
\begin{equation}
  \tau = \frac{n_i}{2 d n_s}
  \sqrt{\frac{\log(\rho(\ve{H}))}{\log(W_c)}}\:.
  \label{eq:optical_thickness_3}
\end{equation}

Using this optical thickness, we can then complete the leakage
approximations by defining the bounds of $\tau \rightarrow 0, \Lambda
\rightarrow 1$ and $\tau \rightarrow \infty, \Lambda \rightarrow
\tau^{-1}$. We expect this behavior as optically thin domains will
communicate most of their histories while optically thick domains will
leak the fraction of histories that did not interact within. With
these bounds, we can then define the leakage fraction out of a domain
for the adjoint Neumann-Ulam method using the Wigner rational
approximation:
\begin{equation}
  \Lambda = \frac{1}{1+\tau}\:,
  \label{eq:wigner_domain_leakage}
\end{equation}
and using the mean-chord approximation:
\begin{equation}
  \Lambda = \frac{1-e^{-\tau}}{\tau}\:.
  \label{eq:mean_chord_domain_leakage}
\end{equation}

Here, the leakage fraction is explicitly bound to the Eigenvalues of
the iteration matrix, the size of the domain, the content of the
discretization stencil, and the weight cutoff selected to terminate
low weight histories.  We also note here an interesting consequence of
this approximation: no matter how optically thick a domain may be, the
domain is still expected to leak some histories and thus communicate
them. Therefore, no matter how large the domains are made, if they do
not encapsulate the entire global domain they require some
communication in order to implement a domain decomposed adjoint
Neumann-Ulam method.

%%---------------------------------------------------------------------------%%
\section{Numerical Experiments}
To test the relationships developed by the spectral analysis, we form
two simple numerical experiments using the diffusion model problem:
one to measure the length of the random walks as a function of the
iteration matrix Eigenvalues, and one to measure the domain leakage
fraction as a function of the iteration matrix Eigenvalues and the
discretization properties. Before doing this, we verify our
computation of the spectral radius of the iteration matrix by
numerically computing the largest Eigenvalue of the diffusion operator
using an iterative Eigenvalue solver. For this verification, a $100
\times 100$ square grid with $h=0.1$ and non-reentrant current
boundary conditions was used and the absorption cross varied from 0 to
100 while the scattering cross section was fixed at
unity. Figure~\ref{fig:measured_spec_rad} gives the measured spectral
radius of the iteration matrix and the computed spectral radius for
the preconditioned diffusion operator using
Eq~(\ref{eq:iteration_radius}) as function of the absorption to
scattering ratio $(\Sigma_a / \Sigma_s)$. We note here that although
there is excellent agreement between the measured and analytic
results, slight differences arise due to the non-reentrant
current condition not being included in the spectral analysis.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in,clip]{measured_spec_rad.png}
  \end{center}
  \caption{\textbf{Measured and analytic preconditioned diffusion
      operator spectral radius as a function of the absorption cross
      section to scattering cross section ratio.} \textit{Values of
      $h=1$, $h=0.1$, and $h=0.01$ were used to verify the
      behavior. The red data was computed numerically by an
      Eigensolver while the black dashed data was generated by
      Eq~(\ref{eq:iteration_radius}).}}
  \label{fig:measured_spec_rad}
\end{figure}

\subsection{Random Walk Length}
With the spectral data in hand, we can go about setting up an
experiment to measure the length of the random walks generated by the
adjoint Neumann-Ulam solver. To do this, we again use a $100 \times
100$ square grid with $h=0.1$ and non-reentrant current boundary
conditions was used and the absorption cross varied from 0 to 100
while the scattering cross section was fixed at unity. Three weight
cutoff values of \sn{1}{-2}, \sn{1}{-4}, and \sn{1}{-8} were used with
10,000 histories generated by a point source source of strength 1 in
the center of the domain such that the vacuum boundary conditions will
have less of an effect on the length of the random walks. For each of
the histories, the number of transitions made was tallied to provide
an effective value of $k$ for each history. This value was then
averaged over all histories to get a measured value of $k$ for the
particular operator. Figure~\ref{fig:measured_length} gives these
measurements as well as the analytic result computed by
Eq~(\ref{eq:analytic_k}) as a function of the iteration matrix
spectral radius, $\rho(\ve{H})$. Again, we note here that although
there is excellent agreement between the measured and analytic
results, slight differences arise due to the non-reentrant current
condition not being included in the spectral analysis.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in,clip]{measured_length.png}
  \end{center}
  \caption{\textbf{Measured and analytic random walk length as a
      function of the iteration matrix spectral radius.} \textit{The
      weight cutoff was varied with \sn{1}{-2}, \sn{1}{-4}, and
      \sn{1}{-8}, to verify the validity of that parameter in
      Eq~(\ref{eq:analytic_k}). The red data was computed numerically
      by an adjoint Neumann-Ulam implementation while the black dashed
      data was generated by Eq~(\ref{eq:analytic_k}).}}
  \label{fig:measured_length}
\end{figure}

\subsection{Domain Leakage}
Finally, we seek to measure the leakage from a domain in a domain
decomposed Monte Carlo calculation and assess the quality of our
analytic relation for the optical thickness of domain and the
associated Wigner approximation. For this experiment, we again use a
$100 \times 100$ square grid with $h=0.1$ and non-reentrant current
boundary conditions. The grid is then decomposed into 100 square
domains, 10 in each cardinal direction. For the cross sections, the
absorption cross was varied from 1 to 100 while the scattering cross
section was set to zero to create a purely absorbing environment. The
optical thickness of these domains will vary as a function of the
absorption cross section if the other parameters are fixed. To compute
the optical thickness, along with the spectral radius as given by
Eq~(\ref{eq:iteration_radius}), we also need the parameters $n_i$ and
$n_s$ which respectively describe the typical domain length and the
average number of states moved along that typical length per history
transition. For our grid above, the domains are of size $10 \times 10$
cells and therefore choosing the $i$ direction there are $n_i=10$ grid
points or states along the typical length of the domain. Looking at
the Laplacian stencil in Figure~\ref{fig:stencil}, we see that all
history transitions will only move a single state in either the $i$ or
$j$ directions due to the symmetry of the problem. We therefore will
move on average 1 state along the typical length of the domain per
history transition, giving $n_s=1$. Figure~\ref{fig:optical_thickness}
gives the domain optical thickness as a function of $\Sigma_a$ as
specified by Eq~(\ref{eq:matrix_optical_thickness_2}) using the above
parameters. Here we note that as the domains become more optically
thick, this relationship becomes exponential as expected.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in,clip]{optical_thickness.png}
  \end{center}
  \caption{\textbf{Analytic domain optical thickness as a function of
      the absorption cross section.} \textit{The data was
      generated by Eq~(\ref{eq:matrix_optical_thickness_2}).}}
  \label{fig:optical_thickness}
\end{figure}

To compute the leakage fraction numerically, over the global domain
\sn{1}{5} histories were sampled from a uniform source of strength
unity. At the start of a stage of histories, the number of histories
starting in the domain was computed and as the stage progressed, the
number of histories that exited the domain was tallied with the ratio
of the two numbers providing a numerical measure for the leakage
fraction. Figure~\ref{fig:measured_length} gives the domain leakage
measurements for the domain in the center of the global grid as well
as the analytic result computed by
Eq~(\ref{eq:analytic_domain_leakage}) as a function of the absorption
cross section. Again, we note good qualitative agreement between the
measured and analytic quantities.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in,clip]{leakage_variation.png}
  \end{center}
  \caption{\textbf{Measured and analytic domain leakage as a function
      of the iteration matrix spectral radius.} \textit{To test the
      behavior with respect to domain size, $n_i=50$, $n_i=100$,and
      $n_i=200$ were used. The red data was computed numerically by a
      domain-decomposed adjoint Neumann-Ulam implementation, the black
      fine-dashed data was generated by
      Eq~(\ref{eq:mean_chord_domain_leakage}), and the dashed-dotted
      black data was generated by
      Eq~(\ref{eq:wigner_domain_leakage}).}}
  \label{fig:measured_leakage}
\end{figure}

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in,clip]{leakage_error.png}
  \end{center}
  \caption{\textbf{Measured and analytic domain leakage absolute error
      as a function of the iteration matrix spectral radius.}
    \textit{To test the behavior with respect to domain size, $n_i=5$
      (blue), $n_i=10$ (red), and $n_i=100$ (yellow) were used. The
      dashed lines represent the error using the Wigner approximation
      while the solid lines represent the error using the mean-chord
      approximation.}}
  \label{fig:measured_leakage}
\end{figure}

>>>>

\subsection{Domain Leakage Approximations}
In a domain decomposed situation, not all histories will remain within
the domain they started in and must instead be communicated. This
communication, expected to be expensive, was analyzed by Siegel and
colleagues for idealized, load balanced situations for full nuclear
reactor core Monte Carlo simulations \cite{siegel_analysis_2012}.  To
quantify the number of particles that leak out of the local domain
they define a leakage fraction, $\Lambda$, as:
\begin{equation}
  \Lambda = \frac{average\ \#\ of\ particles\ leaving\ local\ domain}
          {total\ of\ \#\ of\ particles\ starting\ in\ local\ domain}\:.
  \label{eq:leakage_fraction}
\end{equation}
For their studies, they assumed that the value of $\Lambda$ was
dependent on the total cross section of the system via the Wigner
rational approximation. This approximation was developed by Wigner to
compute the neutron leakage probability from the fuel in a reactor as
a simple interpolation between the small and large fuel lump cases,
approximately characterizing the leakage behavior of many common
geometries \cite{duderstadt_nuclear_1976}.

@@

In the case of domain decomposed linear operator equations, we can use
diffusion theory to estimate the optical thickness of a domain in the
decomposition and the corresponding leakage fraction in terms of
properties of the linear operator and the discretization. To begin we
must first calculate the mean distance a Monte Carlo history will move
in the grid by computing the mean squared distance of its movement
along the chord defined across the domain of length $l$. After a
single transition a history will have moved a mean squared distance
of:
\begin{equation}
  \langle \bar{r_1^2} \rangle = (n_s h)^2\:,
  \label{eq:step_1_length}
\end{equation}
where $h$ is the size of the grid along the chord and $n_s$ is the
number of grid elements a history will move on average every
transition. For our diffusion model problem, $n_s$ would equate to the
number of states in the $i$ (or $j$ as the problem is symmetric)
direction that a history could potentially move in a single transition
and is dependent on the stencil used for the discretization. After $k$
transitions in the random walk, the history will have moved a mean
squared distance of:
\begin{equation}
  \langle \bar{r_k^2} \rangle = k (n_s h)^2\:.
  \label{eq:step_k_length}
\end{equation}
If our our chord is of length $l$ and there are $n_i$ grid points (or
states to which a history may transition) along that chord, then $h =
l / n_i$ giving:
\begin{equation}
  \langle \bar{r_k^2} \rangle = k \Bigg(\frac{n_s l}{n_i}\Bigg)^2\:.
  \label{eq:step_k_length_sub}
\end{equation}
From diffusion theory, we expect the average number of interactions
along the chord to be:
\begin{equation}
  \tau = \frac{l}{2 d \sqrt{\langle \bar{r_k^2} \rangle}}\:,
  \label{eq:optical_thickness_1}
\end{equation}
where $d$ is the dimension of the problem and $\sqrt{\langle
  \bar{r_k^2} \rangle}$ is effectively the mean free path of the Monte
Carlo history in the domain. We can readily interperet $\tau$ to be
the \textit{effective optical thickness} of the domain of length
$l$. Inserting Eq~(\ref{eq:step_k_length_sub}) we arrive at:
\begin{equation}
  \tau = \frac{n_i}{2 d n_s \sqrt{k}}\:,
  \label{eq:optical_thickness_2}
\end{equation}
which if expanded with Eq~(\ref{eq:analytic_k}) gives us the final
relation for the effective optical thickness:
\begin{equation}
  \tau = \frac{n_i}{2 d n_s}
  \sqrt{\frac{\log(\rho(\ve{H}))}{\log(W_c)}}\:.
  \label{eq:optical_thickness_3}
\end{equation}

Using this optical thickness, we can then complete the leakage
approximations by defining the bounds of $\tau \rightarrow 0, \Lambda
\rightarrow 1$ and $\tau \rightarrow \infty, \Lambda \rightarrow
\tau^{-1}$. We expect this behavior as optically thin domains will
communicate most of their histories while optically thick domains will
leak the fraction of histories that did not interact within. With
these bounds, we can then define the leakage fraction out of a domain
for the adjoint Neumann-Ulam method using the Wigner rational
approximation:
\begin{equation}
  \Lambda = \frac{1}{1+\tau}\:,
  \label{eq:wigner_domain_leakage}
\end{equation}
and using the mean-chord approximation:
\begin{equation}
  \Lambda = \frac{1-e^{-\tau}}{\tau}\:.
  \label{eq:mean_chord_domain_leakage}
\end{equation}
Here, the leakage fraction is explicitly bound to the Eigenvalues of
the iteration matrix, the size of the domain, the content of the
discretization stencil, and the weight cutoff selected to terminate
low weight histories.  We also note here an interesting consequence of
this approximation: no matter how optically thick a domain may be, the
domain is still expected to leak some histories and thus communicate
them. Therefore, no matter how large the domains are made, if they do
not encapsulate the entire global domain they require some
communication in order to implement a domain decomposed adjoint
Neumann-Ulam method.

%%---------------------------------------------------------------------------%%
\section{Numerical Experiments}
To test the relationships developed by the spectral analysis, we form
two simple numerical experiments using the diffusion model problem:
one to measure the length of the random walks as a function of the
iteration matrix Eigenvalues, and one to measure the domain leakage
fraction as a function of the iteration matrix Eigenvalues and the
discretization properties. Before doing this, we verify our
computation of the spectral radius of the iteration matrix by
numerically computing the largest Eigenvalue of the diffusion operator
using an iterative Eigenvalue solver. For this verification, a $100
\times 100$ square grid with $h=0.01$, $h=0.1$, and $h=1.0$ and
non-reentrant current boundary conditions was used and the absorption
cross varied from 0 to 100 while the scattering cross section was
fixed at unity. Figure~\ref{fig:measured_spec_rad} gives the measured
spectral radius of the iteration matrix and the computed spectral
radius for the preconditioned diffusion operator using
Eq~(\ref{eq:iteration_radius}) as function of the absorption to
scattering ratio $(\Sigma_a / \Sigma_s)$. Excellent agreement was
observed between the analytic and numerical results with all data
points computed within the tolerance of the iterative Eigenvalue
solver.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in,clip]{measured_spec_rad.png}
  \end{center}
  \caption{\textbf{Measured and analytic preconditioned diffusion
      operator spectral radius as a function of the absorption cross
      section to scattering cross section ratio.} \textit{Values of
      $h=0.01$, $h=0.1$, and $h=1.0$ were used to verify the
      behavior. The red data was computed numerically by an
      Eigensolver while the black dashed data was generated by
      Eq~(\ref{eq:iteration_radius}).}}
  \label{fig:measured_spec_rad}
\end{figure}

\subsection{Random Walk Length}
With the spectral data in hand, we can go about setting up an
experiment to measure the length of the random walks generated by the
adjoint Neumann-Ulam solver. To do this, we again use a $100 \times
100$ square grid with $h=0.1$ and non-reentrant current boundary
conditions was used and the absorption cross varied from 0 to 100
while the scattering cross section was fixed at unity. Three weight
cutoff values of \sn{1}{-2}, \sn{1}{-4}, and \sn{1}{-8} were used with
10,000 histories generated by a point source source of strength 1 in
the center of the domain such that the vacuum boundary conditions will
have less of an effect on the length of the random walks. For each of
the histories, the number of transitions made was tallied to provide
an effective value of $k$ for each history. This value was then
averaged over all histories to get a measured value of $k$ for the
particular operator. Figure~\ref{fig:measured_length} gives these
measurements as well as the analytic result computed by
Eq~(\ref{eq:analytic_k}) as a function of the iteration matrix
spectral radius, $\rho(\ve{H})$. Again, we note good agreement between
the measured and analytic results.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in,clip]{measured_length.png}
  \end{center}
  \caption{\textbf{Measured and analytic random walk length as a
      function of the iteration matrix spectral radius.} \textit{The
      weight cutoff was varied with \sn{1}{-2}, \sn{1}{-4}, and
      \sn{1}{-8}, to verify the validity of that parameter in
      Eq~(\ref{eq:analytic_k}). The red data was computed numerically
      by an adjoint Neumann-Ulam implementation while the black dashed
      data was generated by Eq~(\ref{eq:analytic_k}).}}
  \label{fig:measured_length}
\end{figure}

\subsection{Domain Leakage}
Finally, we seek to measure the leakage from a domain in a domain
decomposed Monte Carlo calculation and assess the quality of our
analytic relation for the optical thickness of domain and the
associated leakage approximations. For this experiment, square grid
with $h=0.1$ and non-reentrant current boundary conditions. The grid
is then decomposed into 9 square domains, 3 in each cardinal
direction. For the cross sections, the absorption cross was varied
from 1 to 100 while the scattering cross section was set to zero to
create a purely absorbing environment. The optical thickness of these
domains will vary as a function of the absorption cross section if the
other parameters are fixed. To compute the optical thickness, along
with the spectral radius as given by Eq~(\ref{eq:iteration_radius}),
we also need the parameters $n_i$ and $n_s$ which respectively
describe the typical domain length and the average number of states
moved along that typical length per history transition. For our grid
above, the domains are varied in size with $50 \times 50$, $100 \times
100$, and $200 \times 200$ cells giving $n_i=50$, $n_i=100$, and
$n_i=200$ grid points or states along the typical length of the domain
respectively. Looking at the Laplacian stencil in
Figure~\ref{fig:stencil}, we see that all history transitions will
only move a single state in either the $i$ or $j$ directions due to
the symmetry of the problem. Furthermore, if we choose the $i$
direction, not all states we will transition to will move the history
in that direction. Therefore, we look to the definition of the
iteration matrix in Eq~(\ref{eq:iteration_stencil}) and the definition
of the adjoint probability matrix in Eq~(\ref{eq:adjoint_probability})
to estimate the $n_s$ parameter. For a particular transition starting
at state $(i,j)$, 6 of the 8 possible new states in the stencil move
the history in $i$ direction with
giving $n_i = \frac{3}{5}$.

To compute the leakage fraction numerically, over the global domain
\sn{1}{5} histories were sampled from a uniform source of strength
unity. At the start of a stage of histories, the number of histories
starting in the domain was computed and as the stage progressed, the
number of histories that exited the domain was tallied with the ratio
of the two numbers providing a numerical measure for the leakage
fraction. Figure~\ref{fig:measured_length} gives the domain leakage
measurements for the domain in the center of the global grid as well
as the analytic result computed by
Eq~(\ref{eq:analytic_domain_leakage}) as a function of the absorption
cross section. Again, we note good qualitative agreement between the
measured and analytic quantities.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in,clip]{leakage_variation.png}
  \end{center}
  \caption{\textbf{Measured and analytic domain leakage as a function
      of the iteration matrix spectral radius.} \textit{To test the
      behavior with respect to domain size, $n_i=50$, $n_i=100$,and
      $n_i=200$ were used. The red data was computed numerically by a
      domain-decomposed adjoint Neumann-Ulam implementation, the black
      fine-dashed data was generated by
      Eq~(\ref{eq:mean_chord_domain_leakage}) using the mean-chord
      approximation, and the dashed-dotted black data was generated by
      Eq~(\ref{eq:wigner_domain_leakage}) using the Wigner rational
      approximation.}}
  \label{fig:measured_leakage}
\end{figure}

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in,clip]{leakage_error.png}
  \end{center}
  \caption{\textbf{Measured and analytic domain leakage absolute error
      as a function of the iteration matrix spectral radius.}
    \textit{To test the behavior with respect to domain size, $n_i=50$
      (blue), $n_i=100$ (red), and $n_i=200$ (yellow) were used. The
      dashed lines represent the error using the Wigner rational
      approximation while the solid lines represent the error using
      the mean-chord approximation.}}
  \label{fig:measured_leakage}
\end{figure}
